\chapter{}

\section*{Rings}

We will devote five lectures to basic theory of rings. The topics to cover are:
1) Basic definitions and examples, 2) ideals, homomorphisms and quotient rings,
3) Chinese remainder theorem, 
4) noetherian rings and Hilbert's theorem, 
5) factorization in commutative rings, 
6) Fermat's theorem, and 
7) Zorn's lemma and maximal ideals. 

\begin{definition}
\index{Ring}
A \textbf{ring} is a set $R$ with two binary operations, the addition
$R\times R\to R$, $(x,y)\mapsto x+y$, and the multiplication
$R\times R\to R$, $(x,y)\mapsto xy$, such that
the following properties hold:
\begin{enumerate}
    \item $(R,+)$ is an abelian group.
    \item $(xy)z=x(yz)$ for all $x,y,z\in R$.
    \item $x(y+z)=xy+xz$ for all $x,y,z\in R$.
    \item $(x+y)z=xz+yz$ for all $x,y,z\in R$.
    \item There exists $1_R\in R$ such that $x1_R=1_Rx=x$ for all $x\in R$.
\end{enumerate}
\end{definition}

Our definition of a ring is that of a ring with identity. In general one
writes the identity element $1_R$ as $1$ if there is no risk of confusion.

\begin{definition}
\index{Ring!commutative}
A ring $R$ is said to be \textbf{commutative} if $xy=yx$ for all $x,y\in R$. 
\end{definition}

\begin{example}
$\Z$, $\Q$, $\R$ and $\C$ are commutative rings.
\end{example}

\begin{example}
    The set  
    \[
    \R[X]=\left\{\sum_{i=0}^na_iX^i:n\in\N_0,\,a_1,\dots,a_n\in \R\right\}
    \]
    of real polynomials in one variable 
    is a commutative ring with the usual operations. 
\end{example}

More generally, if $R$ is a commutative ring, then $R[X]$ is a commutative ring. This construction
allows us to define 
the polynomial ring $R[X,Y]$ in two commuting variables $X$ and $Y$ and coefficients in $R$ as 
$R[X,Y]=(R[X])[Y]$. One can also define the ring  
$R[X_1,\dots,X_n]$ of real polynomials 
in $n$ commuting variables $X_1,\dots,X_n$ with coefficients in $R$ as 
$R[X_1,\dots,X_n]
=(R[X_1,\dots,X_{n-1}])[X_n]$.

\begin{example}
    If $A$ is an abelian group, then 
    the set 
    $\End(A)$ of group homomorphisms $A\to A$ is a ring with
    \[
    (f+g)(x)=f(x)+g(x),\quad
    (fg)(x)=f(g(x)),\quad f,g\in\End(A)\text{ and }x\in A.
    \]
\end{example}

Let $R$ be a ring. 
Some facts:
\begin{enumerate}
    \item $x0=0x=x$ for all $x\in R$.
    \item $x(-y)=-xy$ for all $x,y\in R$.
    \item If $1=0$, then $|R|=1$. 
\end{enumerate}

\begin{example}
    The real vector space $H(\R)=\{a1+bi+cj+dk:a,b,c,d\in\R\}$ with basis $\{1,i,j,k\}$ 
    is a ring with the multiplication induced by
    the formulas 
    \[
    i^2=j^2=k^2=-1,
    \quad ij=k,
    \quad jk=i,
    \quad ki=j.
    \]
    As an example, let us perform a calculation in $H(\R)$: 
    \[
    (1+i+j)(i+k)=i+k-1+ik+ji+jk=i+k-1-j-k+i=-1+2i-j,
    \]
    as $ij=i(ij)=-j$. This is the ring of real \textbf{quaternions}.
\end{example}

\begin{example}
    Let $n\geq2$. 
    The abelian group $\Z/n=\{0,1,\dots,n\}$ of integers modulo $n$ is a ring 
    with the usual multiplication modulo $n$. 
\end{example}

\begin{example}
    Let $n\geq1$. 
    The set $M_n(\R)$ of real $n\times n$ matrices is a ring with the usual matrix operations. Recall
    that if $a=(a_{ij})$ and $b=(b_{ij})$, the multiplication $ab$ is given by
    \[
    (ab)_{ij}=\sum_{k=1}^n a_{ik}b_{kj}.
    \]
\end{example}

Similarly, for any ring $R$ one defines the ring $M_n(R)$ of $n\times n$ matrices
with coefficients in $R$. 

\begin{definition}
\index{Subring}
    Let $R$ be a ring. A \textbf{subring} $S$ of $R$ is a subset $S$ such that
    $(S,+)$ is a subgroup of $(R,+)$ such that $1\in S$ and 
    if $x,y\in S$, then $xy\in S$. 
\end{definition}

Clearly, $\Z\subseteq\Q\subseteq\R\subseteq\C$ is a chain of subrings. 

\begin{example}
    \index{Gauss integers}
    $\Z[i]=\{a+bi:a,b\in\Z\}$ is a subring of $\C$. 
    This is known as the ring of \textbf{Gauss integers}.  
\end{example}

\begin{example}
    $\Q[\sqrt{2}]=\{a+b\sqrt{2}:a,b\in\Q\}$ is a subring of $\R$. 
\end{example}

\begin{example}
    \index{Center!of ring}
    If $R$ is a ring, then the \textbf{center} 
    $Z(R)=\{x\in R:xy=yx\text{ for all $y\in R$}\}$ 
    is a subring of $R$. 
\end{example}

If $S$ is a subring of a ring $R$, then the zero element 
of $S$ is the zero element of $R$, i.e. $0_R=0_S$. Moreover, 
the additive inverse of an element $s\in S$ 
is the additive inverse of $s$ as an element of $R$. 	

\begin{exercise}\
\begin{enumerate}
	\item If $S$ and $T$ are subrings of $R$, then $S\cap T$ is a subring of $R$.
	\item If $R_1\subseteq R_2\subseteq\cdots$ is a sequence of subrings of $R$, then 
	$\cup_{i\geq1}R_i$ is a subring of $R$. 
\end{enumerate}
\end{exercise}

\begin{definition}
\index{Units}
	Let $R$ be a ring. An element $x\in R$ is 
	a \textbf{unit} if there exists $y\in R$ such that $xy=yx=1$. 
\end{definition}

The set $\mathcal{U}(R)$ of units of a ring $R$ form 
a group with the multiplication. For example,
$\mathcal{U}(\Z/8)=\{1,3,5,7\}$. 

\begin{definition}
	\index{Ring!division}
	A \textbf{division ring} is a ring $R$ 
	such that $\mathcal{U}(R)=R\setminus\{0\}$.  	
\end{definition}

The ring $H(\R)$ real quaternions is a non-commutative division ring. Find the inverse of
an arbitrary element $a1+bi+cj+dk\in H(\R)$. 

\begin{definition}
\index{Field}
	A \textbf{field} is a commutative division ring with $1\ne 0$. 
\end{definition}

Clearly, $\Q$, $\R$ and $\C$ are fields. 
If $p$ is a prime number, then $\Z/p$ is a field. 	

\begin{exercise}
	Prove that $\Q[\sqrt{2}]$ is a field. 
	Find the multiplicative inverse of a non-zero element of the form 
	$x+y\sqrt{2}\in\Q[\sqrt{2}]$.  
\end{exercise}

More challenging: Prove that 
\[
\Q[\sqrt[3]{2}]=\{x+y\sqrt[3]{2}+z\sqrt[3]{4}:x,y,z\in\Q\}
\]
is a field. What is the inverse of a non-zero element of the form $x+y\sqrt[3]{2}+z\sqrt[3]{4}$?

\begin{definition}
\index{Ideal!left}
	Let $R$ be a ring. A \textbf{left ideal} of $R$ is a subset $I$ such that 
	$(I,+)$ is a subgroup of $(R,+)$ and such that $RI\subseteq I$, 
	i.e. $ry\in I$ for all $r\in R$ and $y\in I$. 
\end{definition}

Similarly one defines right ideals, one needs 
to replace the condition $RI\subseteq I$ by 
the inclusion 
$IR\subseteq I$. 

\begin{example}
Let $R=M_{2}(\mathbb{R})$. Then   
\[
I=\left(\begin{array}{cc}
\mathbb{R} & \mathbb{R}\\
0 & 0
\end{array}\right)=\left\{ \left(\begin{array}{cc}
x & y\\
0 & 0
\end{array}\right):x,y\in\mathbb{R}\right\} 
\]
is a right ideal $R$ that is not a left ideal. 
\end{example}

Can you find an 
example of a right ideal that is not a left ideal?

\begin{definition}
\index{Ideal}	
Let $R$ be a ring. An ideal of $R$ is a subset that is both a left and a right ideal of $R$. 
\end{definition}
 
If $R$ is a ring, then $\{0\}$ and $R$ are both ideals of $R$. 

\begin{exercise}
Let $R$ be a ring. 
\begin{enumerate}
\item If $\{I_\alpha:\alpha\}$ is a collection of ideals of $R$, then $\cap_{\alpha}I_\alpha$ is an ideal of $R$.  	
\item If $I_1\subseteq I_2\subseteq\cdots$ is a sequence of ideals of $R$, then $\cup_{i\geq1}I_i$ is an ideal of $R$. 
\end{enumerate}
\end{exercise}

\begin{example}
Let $R=\R[X]$. If $f(X)\in R$, then the set 
\[
(f(X))=\{f(X)g(X):g(X)\in R\}
\]
of multiples of $f(X)$ is an ideal of $R$. One can prove that this is the smallest 
ideal of $R$ containing $f(X)$.  	
\end{example}

If $R$ is a ring and $X$ is a subset of $R$, one defines
the ideal generated by $X$ as the smallest ideal of $R$ containing $X$, that is 
\[
(X)=\bigcap\{I:\text{$I$ ideal of $R$ such that $X\subseteq I$}\}.
\]
One proves that 
\[
(X)=\left\{\sum_{i=1}^mr_ix_is_i:m\in\N_0,\,r_1,\dots,r_m,s_1,\dots,s_m\in R\right\}, 
\]
where by convention the empty sum is equal to zero. If $X=\{x_1,\dots,x_n\}$ is a finite
set, then we write $(X)=(x_1,\dots,x_n)$. 

\begin{exercise}
\label{xca:ideals_Z}
Prove that every ideal of $\Z$ is of the form $n\Z$ for some $n\geq0$. 	
\end{exercise}

\begin{exercise}
Let $n\geq2$. Find the ideals of $\Z/n$. 	
\end{exercise}

\begin{exercise}
Find the ideals of $\R$.	
\end{exercise}

A similar exercise is to find the ideals of any division ring.  

\begin{definition}
Let $R$ be a ring and $I$ be an ideal of $R$. Then $I$ is \textbf{principal}
if $I=(x)$ for some $x\in R$. 	
\end{definition}

The division algorithm shows that every ideal of $\Z$ is principal, 
see Exercise~\ref{xca:ideals_Z}.  

\begin{exercise}
	Prove that every ideal of $\R[X]$ is principal. 
\end{exercise}

If $K$ is a field, there is a division algorithm in the 
polynomial ring $K[X]$. Then one proves 
that every ideal of $K[X]$ is principal.  

\begin{exercise}
	Let $R$ be a ring and $x\in R$. Prove that $x\in\mathcal{U}(R)$ if and only if
	$(x)=R$. 	
\end{exercise}

A division ring (and, in particular, a field) has only two ideals. 

\begin{definition}
\index{Ring!homomorphism}
Let $R$ and $S$ be rings. A map $f\colon R\to S$ is a \textbf{ring homomorphism}  
if $f(1)=1$, $f(x+y)=f(x)+f(y)$ and $f(xy)=f(x)f(y)$ for all $x,y\in R$. 	
\end{definition}

Our definition of a ring is that of a ring with identity. This means
that the identity element $1$ of a ring $R$ 
is part of the structure. For that reason, in 
the definition
of a ring homomorphism $f$ one needs $f(1)=1$.  

\begin{example}
The map $f\colon\Z/6\to\Z/6$, $x\mapsto 3x$, is not a ring homomorphism because
$f(1)=3$. 	
\end{example}
 
If $R$ is a ring, then  
the identity map $\id\colon R\to R$, $x\mapsto x$, is a ring homomorphism. 	

\begin{example}
The inclusions $\Z\hookrightarrow\Q\hookrightarrow\R\hookrightarrow\C$ 
are ring homomorphisms. 	
\end{example}

More generally, if $S$ is a subring of a ring $R$, then the inclusion map 
$S\hookrightarrow R$ is a ring homomorphism. 

\begin{example}
Let $R$ be a ring. 
The map $\Z\to R$, $k\mapsto k1$, is a ring homomorphism. 	
\end{example}

\begin{example}
Let $x_0\in\R$. The evaluation map $\R[X]\to\R$, $f\mapsto f(x_0)$, 
is a ring homomorphism. 	
\end{example}

The \textbf{kernel} of a ring homomorphism
$f\colon R\to S$ is the subset
\[
\ker f=\{x\in R:f(x)=0\}.
\]
One proves that the kernel of $f$ is an ideal of $R$. 
Moreover, $\ker f=\{0\}$ if and only if $f$ is injective. The image 
\[
f(R)=\{f(x):x\in R\}
\]
is a subring of $S$. In general, $f(R)$ is not an ideal of $S$. 

\begin{example}
	The map $\C\to M_2(\R)$, $a+bi\mapsto\begin{pmatrix}a&b\\-b&a\end{pmatrix}$, is an injective
	ring homomorphism. 	
\end{example}

\begin{example}
The map $\Z[i]\to\Z/5$, $a+bi\mapsto a+2b\bmod 5$, is a ring homomorphism 
with $\ker f=\{a+bi:a+2b\equiv 0\bmod 5\}$. 	
\end{example}

\begin{exercise}
There is no ring homomorphism $\Z/6\to\Z/15$. Why?	
\end{exercise}

\begin{exercise}
If $f\colon\R[X]\to\R$ is a ring homomorphism 
such that the restriction $f|_{\R}$ of 
$f$ onto $\R$ is the identity, then there exists $x_0\in\R$ such that 
$f$ is the evaluation map at $x_0$. 
\end{exercise}

We now define ring quotients. 
Let $R$ be a ring and $I$ be an ideal of $R$. 
Then $R/I$ is an abelian group
with 
\[
(x+I)+(y+I)=(x+y)+I
\]
and the 
\textbf{canonical map} 
$R\to R/I$, $x\mapsto x+I$,
is a surjective group homomorphism. Recall that 
$R/I$ is the set of cosets $x+I$, where 
$x+I=y+I$ if and only if $x-y\in I$. Note that here we only used
that $I$ is an additive subgroup of $R$. We need an ideal to put a ring structure
on the set $R/I$ of cosets modulo $I$. As in the case of integers, 
we use the following notation. For $x,y\in R$ 
we write 
\[
x\equiv y\bmod I\Longleftrightarrow x-y\in I.
\]

How can we put a ring structure on $R/I$? It makes sense
to define a multiplication on $R/I$ in such a way that
the canonical map $R\to R/I$ is a surjective ring homomorphis. For that purpose, 
we define 
\[
(x+I)(y+I)=(xy)+I.
\]
Since $I$ is an ideal of $R$, this multiplication is well-defined. In fact, let 
$x+I=x_1+I$ and $y+I=y_1+I$. We want to show that
$xy+I=x_1y_1+I$. Since $x-x_1\in I$, 
\[
xy-x_1y=(x-x_1)y\in I
\]
because $I$ is a right ideal. Similarly, since $y-y_1\in I$, it follows that 
\[
x_1y-x_1y_1=x_1(y-y_1)\in I,
\]
as $I$ is a left ideal. Thus
\[
xy-x_1y_1=xy-x_1y+x_1y-x_1y_1=(x-x_1)y+x_1(y-y_1)\in I.
\]

\begin{theorem}
	Let $R$ be a ring and $I$ be an ideal of $R$. Then
	$R/I$ with 
	\[
	(x+I)+(y+I)=(x+y)+I,\quad
	(x+I)(y+I)=(xy)+I,
	\]
	is a ring and the canonical map $R\to R/I$, $x\mapsto x+I$, 
	is a surjective
	ring homomorphism with kernel $I$. 
\end{theorem}

We have already seen that the multiplication is well-defined. The rest
of the proof is left as an exercise. 
