\chapter{Algebras}

\index{Group ring}
\index{Group algebra}
We now discuss an important family of examples. 
Fix a field $K$. 
For a finite group $G$ let $K[G]$ be the vector space (over $K$)
with basis $\{g:g\in G\}$. Then $K[G]$ is a ring
with
\[
\left(\sum_{g\in G}\lambda_gg\right)\left(\sum_{h\in G}\mu_hh\right)
=\sum_{g,h\in G}\lambda_g\mu_h(gh).
\] 

Thus $K[G]$ is a ring and also a vector space (over $K$) and these structures
are somewhat compatible. Note that
\[
(\lambda a+\mu g=c=\lambda (ac)+\mu (bc),\quad
a(\lambda b+\mu c)=\lambda (ab)+\mu (ac)
\]
for all $\lambda,\mu\in K$ and $a,b,c\in K[G]$. 

\begin{definition}
\index{Algebra}
Let $A$ be a ring. Then $A$ is an algebra (over the field $K$) if $A$ is a vector space
and the map $K\to Z(A)$, $k\mapsto k1_A$, is an injective ring homomorphism.  
\end{definition}

Thus $K[G]$ is an algebra, as it is ring that contains $K$ in its center (or more precisely,
the map $K\to Z(K[G])$, $k\mapsto k1$, is an injective ring homomorphism.  
Other examples of algebras 
are the polynomial rings $K[X]$ and $K[X,Y]$ and matrix rings $M_n(K)$.  

\begin{example}
	If $A$ is an algebra, then $M_n(A)$ is an algebra.	
\end{example}

The ring $K[G]$ is commutative if and only if $G$ is abelian. Moreove,
$K[G]$ is a vector space of dimension $\dim K[G]=|G|$.

\begin{example}
	Let $G=\langle g:g^3=1\rangle=\{1,g,g^2\}\simeq C_3$ be the cyclic group of order three. 
	If $\alpha=a_11+_2g+a_3g^2$ and $\beta=b_11+b_2g+b_3g^2$, then
	\[
		\alpha\beta=(a_1b_1+a_2b_3+a_3b_2)1+(a_1b_2+a_2b_1+a_3b_3)g+a_1b_3+a_2b_2+a_3b_1)g^2.
	\]
	One can check that $\C[G]\simeq\C[X]/(X^3-1)$. 
\end{example}

In general, one proves that $\C[C_n]\simeq\C[X]/(X^n-1)$ for $n\geq2$.

\begin{example}
	Let $K$ be a field and 
	$G=\{1,g\}\simeq C_2$ be the cyclic group of order two. The product
	of $K[G]$ is 
	\[
	(a1+bg)(c1+gd)=(ac+bd)1+(ad+bc)g.
	\]
	
	If $K=\C$, then the map $K[G]\to K\times K$, $a1+bg\mapsto (a+b,a-b)$, 
	is a linear isomorphism of rings. 
	
	If $K=\Z/2$, then the map $K[G]\to\begin{pmatrix}
		K&K\\
		0&K
	\end{pmatrix}$, $a1+bg\mapsto\begin{pmatrix}
		a+b&b\\
		0&a+b		
	\end{pmatrix}$, is a linear isomorphism of rings. 
\end{example}

\begin{exercise}
	Prove that $\R[C_3]\simeq\R\times\C$. 	
\end{exercise}

The group ring has the following property, which is left as an exercise. 
zzZLet $R$ be a ring and
$G$ be a finite group. If $f\colon G\to\mathcal{U}(R)$ is a group homomorphism, 
then there exists a unique ring homomorphism $\varphi\colon K[G]\to R$ such that
$\varphi|_G=f$. 

\begin{example}
	Let $\D_3=\langle r,s:r^3=s^2=1,\,srs^{-1}=r^{-1}\rangle$ be the dihedral
	group of six elements. We claim that 
	\[
	\C[\D_3]\simeq\C\times\C\times M_2(\C).
	\]
	Let $\omega$ be a primitive root of one. Let 
	\[
	R=\begin{pmatrix}
		\omega&0\\
		0&\omega^2	
	\end{pmatrix},
	\quad
	S=\begin{pmatrix}
		0&1\\
		1&0
	\end{pmatrix}.
 	\]
 	One easily checks that $SRS^{-1}=R^{-1}$ and $R^3=S^2=\begin{pmatrix}
		1&0\\
		0&1	
	\end{pmatrix}$. It follows that there exists a group homomorphism
	$G\to\C\times\C\times M_2(\C)$ such that
	$r\mapsto (1,1,R)$ and $s\mapsto (1,-1,S)$. This group homomorphism
	is a ring isomorphism.  
\end{example}




 
