\chapter{Rings and ideals}

\begin{definition}
\index{Ring}
A \textbf{ring} is a set $R$ with two binary operations, the addition
$R\times R\to R$, $(x,y)\mapsto x+y$, and the multiplication
$R\times R\to R$, $(x,y)\mapsto xy$, such that
the following properties hold:
\begin{enumerate}
    \item $(R,+)$ is an abelian group.
    \item $(xy)z=x(yz)$ for all $x,y,z\in R$.
    \item $x(y+z)=xy+xz$ for all $x,y,z\in R$.
    \item $(x+y)z=xz+yz$ for all $x,y,z\in R$.
    \item There exists $1_R\in R$ such that $x1_R=1_Rx=x$ for all $x\in R$.
\end{enumerate}
\end{definition}

Our definition of a ring is that of a ring with identity. In general one
writes the identity element $1_R$ as $1$ if there is no risk of confusion.

\begin{definition}
\index{Ring!commutative}
A ring $R$ is said to be \textbf{commutative} if $xy=yx$ for all $x,y\in R$. 
\end{definition}

\begin{example}
$\Z$, $\Q$, $\R$ and $\C$ are commutative rings.
\end{example}

\begin{example}
    The set  
    \[
    \R[X]=\left\{\sum_{i=0}^na_iX^i:n\in\N_0,\,a_1,\dots,a_n\in \R\right\}
    \]
    of real polynomials in one variable 
    is a commutative ring with the usual operations. 
\end{example}

More generally, if $R$ is a commutative ring, then $R[X]$ is a commutative ring. This construction
allows us to define 
the polynomial ring $R[X,Y]$ in two commuting variables $X$ and $Y$ and coefficients in $R$ as 
$R[X,Y]=(R[X])[Y]$. One can also define the ring  
$R[X_1,\dots,X_n]$ of real polynomials 
in $n$ commuting variables $X_1,\dots,X_n$ with coefficients in $R$ as 
$R[X_1,\dots,X_n]
=(R[X_1,\dots,X_{n-1}])[X_n]$.

\begin{example}
    If $A$ is an abelian group, then $\End(A)$ is a ring with
    \[
    (f+g)(x)=f(x)+g(x),\quad
    (fg)(x)=f(g(x)),\quad f,g\in\End(A)\text{ and }x\in A.
    \]
\end{example}

Let $R$ be a ring. 
Some facts:
\begin{enumerate}
    \item $x0=0x=x$ for all $x\in R$.
    \item $x(-y)=-xy$ for all $x,y\in R$.
    \item If $1=0$, then $|R|=1$. 
\end{enumerate}

\begin{example}
    The real vector space $H(\R)=\{a1+bi+cj+dk:a,b,c,d\in\R\}$ with basis $\{1,i,j,k\}$ 
    is a ring with the multiplication induced by
    the formulas 
    \[
    i^2=j^2=k^2=-1,
    \quad ij=k,
    \quad jk=i,
    \quad ki=j.
    \]
    As an example, let us perform a calculation in $H(\R)$: 
    \[
    (1+i+j)(i+k)=i+k-1+ik+ji+jk=i+k-1-j-k+i=-1+2i-j,
    \]
    as $ij=i(ij)=-j$. This is the ring of real \textbf{quaternions}.
\end{example}

\begin{example}
    Let $n\geq2$. 
    The abelian group $\Z/n=\{0,1,\dots,n\}$ of integers modulo $n$ is a ring 
    with the usual multiplication modulo $n$. 
\end{example}

\begin{example}
    Let $n\geq1$. 
    The set $M_n(\R)$ of real $n\times n$ matrices is a ring with the usual matrix operations. Recall
    that if $a=(a_{ij})$ and $b=(b_{ij})$, the multiplication $ab$ is given by
    \[
    (ab)_{ij}=\sum_{k=1}^n a_{ik}b_{kj}.
    \]
\end{example}

Similarly, for any ring $R$ one defines the ring $M_n(R)$ of $n\times n$ matrices
with coefficients in $R$. 

\begin{definition}
\index{Subring}
    Let $R$ be a ring. A \textbf{subring} $S$ of $R$ is a subset $S$ such that
    $(S,+)$ is a subgroup of $(R,+)$ such that $1\in S$ and 
    if $x,y\in S$, then $xy\in S$. 
\end{definition}

Clearly, $\Z\subseteq\Q\subseteq\R\subseteq\C$ is a chain of subrings. 

\begin{example}
    \index{Gauss integers}
    $\Z[i]=\{a+bi:a,b\in\Z\}$ is a subring of $\C$. 
    This is known as the ring of \textbf{Gauss integers}.  
\end{example}

\begin{example}
    $\Q[\sqrt{2}]=\{a+b\sqrt{2}:a,b\in\Q\}$ is a subring of $\R$. 
\end{example}

\begin{example}
    \index{Center!of ring}
    If $R$ is a ring, then the \textbf{center} 
    $Z(R)=\{x\in R:xy=yx\text{ for all $y\in R$}\}$ 
    is a subring of $R$. 
\end{example}

If $S$ is a subring of a ring $R$, then the zero element 
of $S$ is the zero element of $R$, i.e. $0_R=0_S$. Moreover, 
the additive inverse of an element $s\in S$ 
is the additive inverse of $s$ as an element of $R$. 	

\begin{exercise}\
\begin{enumerate}
	\item If $S$ and $T$ are subrings of $R$, then $S\cap T$ is a subring of $R$.
	\item If $R_1\subseteq R_2\subseteq\cdots$ is a sequence of subrings of $R$, then 
	$\cup_{i\geq1}R_i$ is a subring of $R$. 
\end{enumerate}
\end{exercise}

\begin{definition}
\index{Units}
	Let $R$ be a ring. An element $x\in R$ is 
	a \textbf{unit} if there exists $y\in R$ such that $xy=yx=1$. 
\end{definition}

The set $\mathcal{U}(R)$ of units of a ring $R$ form 
a group with the multiplication. For example,
$\mathcal{U}(\Z/8)=\{1,3,5,7\}$. 

\begin{definition}
	\index{Ring!division}
	A \textbf{division ring} is a ring $R$ 
	such that $\mathcal{U}(R)=R\setminus\{0\}$.  	
\end{definition}

The ring $H(\R)$ real quaternions is a non-commutative division ring. Find the inverse of
an arbitrary element $a1+bi+cj+dk\in H(\R)$. 

\begin{definition}
\index{Field}
	A \textbf{field} is a commutative division ring with $1\ne 0$. 
\end{definition}

Clearly, $\Q$, $\R$ and $\C$ are fields. 
If $p$ is a prime number, then $\Z/p$ is a field. 	

\begin{exercise}
	$\Q[\sqrt{2}]$ is a field. 
	Find the multiplicative inverse of $x+y\sqrt{2}\in\Q[\sqrt{2}]$.  
\end{exercise}

More challenging: Prove that 
\[
\Q[\sqrt[3]{2}]=\{x+y\sqrt[3]{2}+z\sqrt[3]{4}:x,y,z\in\Q\}
\]
is a field. What is the inverse of $x+y\sqrt[3]{2}+z\sqrt[3]{4}$?

\begin{definition}
\index{Ideal!left}
	Let $R$ be a ring. A \textbf{left ideal} of $R$ is a subset $I$ such that 
	$(I,+)$ is a subgroup of $(R,+)$ and such that $RI\subseteq I$, 
	i.e. $ry\in I$ for all $r\in R$ and $y\in I$. 
\end{definition}

Similarly one defines right ideals, one needs to replace the condition $RI\subseteq I$ by 
the inclusion 
$IR\subseteq I$. 

\begin{example}
Let $R=M_{2}(\mathbb{R})$. Then   
\[
I=\left(\begin{array}{cc}
\mathbb{R} & \mathbb{R}\\
0 & 0
\end{array}\right)=\left\{ \left(\begin{array}{cc}
x & y\\
0 & 0
\end{array}\right):x,y\in\mathbb{R}\right\} 
\]
is a right ideal $R$ that is not a left ideal. 
\end{example}

Can you find an 
example of a right ideal that is not a left ideal?

\begin{definition}
\index{Ideal}	
Let $R$ be a ring. An ideal of $R$ is a subset that is both a left and a right ideal of $R$. 
\end{definition}
 
If $R$ is a ring, then $\{0\}$ and $R$ are both ideals of $R$. 

\begin{exercise}
Let $R$ be a ring. 
\begin{enumerate}
\item If $\{I_\alpha:\alpha\}$ is a collection of ideals of $R$, then $\cap_{\alpha}I_\alpha$ is an ideal of $R$.  	
\item If $I_1\subseteq I_2\subseteq\cdots$ is a sequence of ideals of $R$, then $\cup_{i\geq1}I_i$ is an ideal of $R$. 
\end{enumerate}
\end{exercise}

\begin{example}
Let $R=\R[X]$. If $f(X)\in R$, then the set 
\[
(f(X))=\{f(X)g(X):g(X)\in R\}
\]
of multiples of $f(X)$ is an ideal of $R$. One can prove that this is the smallest 
ideal of $R$ containing $f(X)$.  	
\end{example}

If $R$ is a ring and $X$ is a subset of $R$, one defines
the ideal generated by $X$ as the smallest ideal of $R$ containing $X$, that is 
\[
(X)=\bigcap\{I:\text{$I$ ideal of $R$ such that $X\subseteq I$}\}.
\]
One proves that 
\[
(X)=\left\{\sum_{i=1}^mr_ix_is_i:m\in\N_0,\,r_1,\dots,r_m,s_1,\dots,s_m\in R\right\}, 
\]
where by convention the empty sum is equal to zero. If $X=\{x_1,\dots,x_n\}$ is a finite
set, then we write $(X)=(x_1,\dots,x_n)$. 

\begin{exercise}
Prove that every ideal of $\Z$ is of the form $n\Z$ for some $n\geq0$. 	
\end{exercise}

\begin{exercise}
Let $n\geq2$. Find the ideals of $\Z/n$. 	
\end{exercise}

\begin{exercise}
Find the ideals of $\R$.	
\end{exercise}

One proves that a field has only two ideals. 

\begin{definition}
\index{Ring!homomorphism}
Let $R$ and $S$ be rings. A map $f\colon R\to S$ is a \textbf{ring homomorphism}  
if $f(1)=1$, $f(x+y)=f(x)+f(y)$ and $f(xy)=f(x)f(y)$ for all $x,y\in R$. 	
\end{definition}

Our definition of a ring is that of a ring with identity. This means
that the identity element $1$ of a ring $R$ 
is part of the structure. For that reason, in 
the definition
of a ring homomorphism $f$ one needs $f(1)=1$.  

\begin{example}
The map $f\colon\Z/6\to\Z/6$, $x\mapsto 3x$, is not a ring homomorphism because
$f(1)=3$. 	
\end{example}
 
If $R$ is a ring, then  
the identity map $\id\colon R\to R$, $x\mapsto x$, is a ring homomorphism. 	

\begin{example}
The inclusions $\Z\hookrightarrow\Q\hookrightarrow\R\hookrightarrow\C$ 
are ring homomorphisms. 	
\end{example}

More generally, if $S$ is a subring of a ring $R$, then the inclusion map 
$S\hookrightarrow R$ is a ring homomorphism. 

\begin{example}
Let $R$ be a ring. 
The map $\Z\to R$, $k\mapsto k1$, is a ring homomorphism. 	
\end{example}

\begin{example}
Let $x_0\in\R$. The evaluation map $\R[X]\to\R$, $f\mapsto f(x_0)$, 
is a ring homomorphism. 	
\end{example}

The \textbf{kernel} of a ring homomorphism
$f\colon R\to S$ is the subset
\[
\ker f=\{x\in R:f(x)=0\}.
\]
One proves that the kernel of $f$ is an ideal of $R$. 
Moreover, $\ker f=\{0\}$ if and only if $f$ is injective. The image 
\[
f(R)=\{f(x):x\in R\}
\]
is a subring of $S$. In general, $f(R)$ is not an ideal of $S$. 

\begin{example}
The map $\C\to M_2(\R)$, $a+bi\mapsto\begin{pmatrix}a&b\\-b&a\end{pmatrix}$, is an injective
ring homomorphism. 	
\end{example}

\begin{example}
The map $\Z[i]\to\Z/5$, $a+bi\mapsto a+2b\bmod 5$, is a ring homomorphism 
with $\ker f=\{a+bi:a+2b\equiv 0\bmod 5\}$. 	
\end{example}

\begin{exercise}
There is no ring homomorphism $\Z/6\to\Z/15$. Why?	
\end{exercise}

\begin{exercise}
If $f\colon\R[X]\to\R$ is a ring homomorphism 
such that the restriction $f|_{\R}$ of 
$f$ onto $\R$ is the identity, then there exists $x_0\in\R$ such that 
$f$ is the evaluation map at $x_0$. 
\end{exercise}

We now define ring quotients. 
Let $R$ be a ring and $I$ be an ideal of $R$. 
Then $R/I$ is an abelian group
with 
\[
(x+I)+(y+I)=(x+y)+I
\]
and the 
\textbf{canonical map} 
$R\to R/I$, $x\mapsto x+I$,
is a surjective group homomorphism. Recall that 
$R/I$ is the set of cosets $x+I$, where 
$x+I=y+I$ if and only if $x-y\in I$. Note that here we only used
that $I$ is an additive subgroup of $R$. We need an ideal to put a ring structure
on the set $R/I$ of cosets modulo $I$. As in the case of integers, 
we use the following notation. For $x,y\in R$ 
we write 
\[
x\equiv y\bmod I\Longleftrightarrow x-y\in I.
\]

How can we put a ring structure on $R/I$? It makes sense
to define a multiplication on $R/I$ in such a way that
the canonical map $R\to R/I$ is a surjective ring homomorphis. For that purpose, 
we define 
\[
(x+I)(y+I)=(xy)+I.
\]
Since $I$ is an ideal of $R$, this multiplication is well-defined. In fact, let 
$x+I=x_1+I$ and $y+I=y_1+I$. We want to show that
$xy+I=x_1y_1+I$. Since $x-x_1\in I$, 
\[
xy-x_1y=(x-x_1)y\in I
\]
because $I$ is a right ideal. Similarly, since $y-y_1\in I$, it follows that 
\[
x_1y-x_1y_1=x_1(y-y_1)\in I,
\]
as $I$ is a left ideal. Thus
\[
xy-x_1y_1=xy-x_1y+x_1y-x_1y_1=(x-x_1)y+x_1(y-y_1)\in I.
\]

\begin{theorem}
	Let $R$ be a ring and $I$ be an ideal of $R$. Then
	$R/I$ with 
	\[
	(x+I)+(y+I)=(x+y)+I,\quad
	(x+I)(y+I)=(xy)+I,
	\]
	is a ring and the canonical map $R\to R/I$, $x\mapsto x+I$, 
	is a surjective
	ring homomorphism with kernel $I$. 
\end{theorem}

We have already seen that the multiplication is well-defined. The rest
of the proof is left as an exercise. 

\begin{example}
	Let $R=(\Z/3)[X]$ and $I=(2X^2+X+2)$ be the ideal of $R$ 
	generated by the polynomial $2X^2+X+2$. 	If $f(X)\in R$, 
	the division algorithm allows us to write
	\[
	f(X)=(2X^2+X+2)q(X)+r(X),
	\]
	for some $q(X),r(X)\in R$, where either $r(X)=0$ or $\deg r(X)<2$. 
	This means
	that $r(X)=aX+b$ for some $a,b\in R$.
	Note that
	$f(X)\equiv aX+b\bmod (2X^2+X+2)$
	for some $a,b\in\Z/3$, 
	so the quotient ring $R/I$ has 
	nine elements.  
\end{example}

As it happens in the case of groups, 
to undertand quotient rings one has the first isomorphism theorem. 

\begin{theorem}[first isomorphism theorem]
	If $f\colon R\to S$ is a ring homomorphism, then $R/\ker f\simeq f(R)$.  	
\end{theorem}

This is somwehat similar to the result one knows from group theory. 
One needs to show that 
the map $R/I\to f(R)$, $x+I\mapsto f(x)$, is a well-defined 
bijective ring homomorphism. 

\begin{example}
	The evaluation map $\R[X]\to\C$, $f(X)\mapsto f(i)$, is a surjective ring
	homomorphism with kernel $(X^2+1)$. Thus $\R[X]/(X^2+1)\simeq\C$ by the
	first isomorphism theorem.   	
\end{example}

\begin{exercise}
	Let $R$ be the ring of continuos maps $[0,2]\to\R$. Prove that
	$I=\{f\in R:f(1)=0\}$ is an ideal of $R$ and that $R/I\simeq\R$.   	
\end{exercise}

\begin{exercise}
Let 
\[
R=\left\{\begin{pmatrix}
a&b\\
0&a
\end{pmatrix}:a,b\in\Q\right\}
\]
Prove that the map $R\to\Q$, $\begin{pmatrix}a&b\\0&a\end{pmatrix}\mapsto a$, is a surjective 
ring homomorphism with kernel 
$I=\left\{\begin{pmatrix}
0&b\\
0&0\end{pmatrix}:b\in\Q\right\}$. 	
Thus $R/I\simeq\Q$. 	
\end{exercise}

\begin{exercise}
	Let $n\geq1$. 
	Let $R$ be a ring and $I$ be an ideal of $R$. Prove that $M_n(I)$ is an ideal 
	of $M_n(R)$ and that $M_n(R)/M_n(I)\simeq M_n(R/I)$. 	
\end{exercise}

\begin{exercise}
	Let $R=\Z[\sqrt{10}]$ and $I=(2,\sqrt{10})$. Prove that $R/I\simeq\Z/2$. 	
\end{exercise}

\begin{exercise}
	Prove that $\Z[i]/(1+3i)\simeq\Z/10$. 	
\end{exercise}

\begin{exercise}
	Prove that there is no ideal $I$ of $\Z[i]$ 
	such that $\Z[i]/I\simeq\Z/15$. 
\end{exercise}

As it happens in group theory, one has the following important result. 

\begin{theorem}[correspondence theorem]
	Let $f\colon R\to S$ be a surjective ring homomorphism. There exists a
	bijective correspondence between 
	the set of ideals of $R$ containing $\ker f$ and
	the set of ideals of $S$.  
\end{theorem}

\begin{proof}[Sketch of proof]
Let $I$ be an ideal of $R$ contaning $\ker f$ and
let $J$ be an ideal of $S$. 
We need to prove the following facts:
\begin{enumerate}
\item $f(I)$ is an ideal of $S$.
\item $f^{-1}(J)$ is an ideal of $R$ containing $\ker f$. 
\item $f(f^{-1}(J))=J$ and $f^{-1}(f(I))=I$. 
\item If $f(I)=J$, then $R/I\simeq S/J$. 
\end{enumerate}
We only prove the fourth statement, the others are left as exercises. Note that
the third claim implies that $f(I)=J$ if and only if $I=f^{-1}(J)$. 
Let 
$\pi\colon S\to S/J$ be the canonical map. The composition
$g=\pi\circ f\colon R\to S/J$ is a ring homomorphism and
\[
\ker g=\{x\in R:g(x)=0\}=\{x\in R:f(x)\in J\}=\{x\in R:x\in f^{-1}(J)=I\}=I.
\]
Since $g(R)=S/J$, the first isomorphism theorem implies that $R/I\simeq S/J$.
\end{proof}


